Seismic Time Series Analysis\\Statistics for Spatio-Temporal Data\\\textit{-- project draft --}

### 2020-06-06 draft.tex
more specific intro and reword/reorder some pieces

@@

@@

Since the periodgrams suggested the presence of two significative periodicities, we also tested the extraction of both corresponding seasonalities:
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item Apply the MA smoothing filter (f=24) on the dataset to extract the trend component
	\item Remove the trend from the dataset and extract the 24-hour seasonal figure using the local mean method
	\item Remove the 24-hour seasonal components to obtain a deseasoned dataset
	\item Apply the MA smoothing filter (f=168) on the deseasoned dataset to extract the updated trend component
	\item Remove the trend from the already deseasoned dataset and extract the 168-hour seasonal figure using the local mean method
	\item Repeat steps 3-5 steps alternating the 168/24 seasonalities until convergence (2 to 3 cycles should be enough)
	\item Removing the last fitted trend and the two seasonalities we obtain the final residuals
\end{enumerate}

@@

In \figurename~\ref{fig:decomposed} we see the decomposition results. While data from Uccle station have a clear significativity of both trend and seasonalities, data from Membach could seem not significant. Still, if we notice that the main timeseries lies between values of 0 and 100, we can find out that around half of the value will be composed by the trend, a 30\% by the joined seasonalities component, and the remaining value by the residuals. We will need to better handle the spikes starting from the middle of our dataset.

Some R commands cannot handle timeseries with missing values, but as we have seen both stations datasets include missing values.
Also, please note that we are going to have a timeseries object indexed by weeks, so that it contains the 168 hours between each integer.
Since we know that the longest missing values series is 90 hours long, we can apply a simple completion strategy for serving those commands a filled timeseries: we will substitute missing values with the mean of the week they are in; in R terms \texttt{na.aggregate(x, floor)}.

@@

The transformed data still shows no constant mean, even if it has a way better shape than the original data. Thus, we consider the usage of the \textit{diff} operator, simple and at lag 168. Interestingly, the analysis of ACF and PACF plots hints to similar models to the Uccle station: 
$\mathrm{ARIMA}(0,1,0)(0,1,1)_{168}$ and $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$.

@@

We can still compare how the seasonalities differ with respect with each original dataset and between each other. We extract the seasonalities for each station highlighting the standard deviation confidence band for an estimated error of 67\%. This estimation lies on the assumption that values for each single hour in the hourly/weekly seasonality is Normally distributed, a slightly different assumption that is not necessarily refuted by the already seen Q-Q plots.

>>>>

The correct estimation of a background noise, correlated or not with human activity, could help institutions to better isolate less prominent earthquakes and similar events from the seismic data. This could result in better understanding of the frequency of earthquakes and by studying their more-precise evolution over time.

@@

Some R commands cannot handle timeseries with missing values, but as we have seen both stations datasets include missing values.
Also, please note that we are going to have a timeseries object indexed by weeks, so that it contains the 168 hours between each integer index.
Since we know that the longest missing values series is 90 hours long, we can apply a simple completion strategy for serving those commands a filled timeseries: we will substitute missing values with the mean of the week they are in; in R terms \texttt{na.aggregate(x, floor)}.

@@

Since the periodgrams suggested the presence of two significative periodicities, we also tested the extraction of both corresponding seasonalities:
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item Apply the MA smoothing filter (f=24) on the dataset to extract the trend component
	\item Remove the trend from the dataset and extract the 24-hour seasonal figure using the local mean method
	\item Remove the 24-hour seasonal components to obtain a deseasoned dataset
	\item Apply the MA smoothing filter (f=168) on the deseasoned dataset to extract the updated trend component
	\item Remove the trend from the already deseasoned dataset and extract the 168-hour seasonal figure using the local mean method
	\item Repeat steps 1-5 using the last deseasoned dataset until convergence (2-3 cycles)
	\item Removing the last fitted trend and the two seasonalities we obtain the final residuals
\end{enumerate}

@@

In \figurename~\ref{fig:decomposed} we see the double decomposition results. While data from Uccle station have a clear significativity of both trend and seasonalities, data from Membach could seem not significant. Still, if we notice that the main timeseries lies between values of 0 and 100, we can find out that around half of the value will be composed by the trend, a 30\% by the joined seasonalities component, and the remaining value by the residuals. We will need to better handle the spikes that start from the middle of that dataset.

@@

The transformed data still shows no constant mean, even if it has a way better shape than the original data, confirmed by the Q-Q plot (\figurename~\ref{fig:mems:box-cox:qq}). Thus, we consider the usage of the \textit{diff} operator, simple and at lag 168. Interestingly, the analysis of ACF and PACF plots hints to similar models to the Uccle station: 
$\mathrm{ARIMA}(0,1,0)(0,1,1)_{168}$ and $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$.

@@

We can still compare how the seasonalities differ with respect with each original dataset and between each other. We extract the seasonalities for each station highlighting the standard deviation confidence band for an estimated error of 67\%. This interval estimation lies on the assumption that values for each single hour in the hourly/weekly seasonality is Normally distributed, a slightly different assumption that is not necessarily refuted by the already seen Q-Q plots.
